E-graphs With Bindings

### 2025-01-09 LICS/sections/appendix.tex
some updates

\begin{align*}
n \xrightarrow{} n' \xrightarrow{} &\;\mathcal{F} \xleftarrow{} m' \xleftarrow{} m\\
&\;+ \hspace{6em} \Coloneqq \hspace{2em} n  \xrightarrow{} n' \xrightarrow{} \;\mathcal{F} \xleftarrow{} m' \xleftarrow{} m\\
0 \xrightarrow{} 0 \xrightarrow{} &\;\varnothing \xleftarrow{} 0 \xleftarrow{} 0.
\end{align*}

@@

\begin{definition}[Hashcons invariant]
The hashcons $H$ must map all canonical e-nodes to their e-class ids:
\[
\text{e-node } n \in M[a] \text{ iff } H[\text{canonicalize}(n)] = \text{find}(a)
\]

\begin{algorithm}
    \caption{Pattern Matching Pseudocode Example}
    \begin{algorithmic}[1]
    \ForAll{e-class-id in H}
    \State edge $edge \gets$ mkEdge(e-class-id)
        \ForAll {e-node in H \textbf{where} H[e-node] == e-class-id}
            \State \textbf{match} e-node \textbf{with}
                \State \quad \textbf{case} \texttt{f}:
                %    \State \quad \quad edge.add(mk_edge(f))
        \EndFor
    \EndFor
    % \Function{matchPattern}{shape}
    %     \State \textbf{match} shape \textbf{with}
    %     \State \quad \textbf{case} \texttt{Circle(radius)}:
    %     \State \quad \quad \Return $\pi \cdot \texttt{radius}^2$
    %     \State \quad \textbf{case} \texttt{Rectangle(width, height)}:
    %     \State \quad \quad \Return \texttt{width} $\times$ \texttt{height}
    %     \State \quad \textbf{case} \texttt{Triangle(base, height)}:
    %     \State \quad \quad \Return $\frac{1}{2} \times \texttt{base} \times \texttt{height}$
    %     \State \quad \textbf{case} \texttt{Square(side)}:
    %     \State \quad \quad \Return \texttt{side}$^2$
    %     \State \quad \textbf{case} \texttt{\_}:
    %     \State \quad \quad \Return ``Unknown shape''
    % \EndFunction
    \end{algorithmic}
    \end{algorithm}
\end{definition}

>>>>

\begin{align*}
0 \xrightarrow{} n \xrightarrow{} &\;\mathcal{F} \xleftarrow{} m \xleftarrow{} 0\\
&\;+ \hspace{6em} \Coloneqq \hspace{2em} 0  \xrightarrow{} n \xrightarrow{} \;\mathcal{F} \xleftarrow{} m \xleftarrow{} 0\\
0 \xrightarrow{} 0 \xrightarrow{} &\;\varnothing \xleftarrow{} 0 \xleftarrow{} 0.
\end{align*}

@@

\begin{definition}[Hashcons invariant]
The hashcons $H$ must map all canonical e-nodes to their e-class ids:
\[
\text{e-node } n \in M[a] \text{ iff } H[\text{canonicalize}(n)] = \text{find}(a)
\]
\end{definition}

The pseudocode below shows the translation from a \textit{canonical} (the one for which the above invariant holds).
For the following e-graph with e-classes shown as dashed boxes it produces the following cospan of e-hypergraphs.
\texttt{connect} and \texttt{mkEdge} are responsible for making things shared using $\triangle$.

\begin{figure}
    \begin{subfigure}{0.4\linewidth}
\[
\adjustbox{scale=0.6}{
\tikzfig{../figures/appendix/e-graph-example_1}
}
\]
\end{subfigure}
\begin{subfigure}{0.4\linewidth}
    \[
    \adjustbox{scale=0.6}{
    \tikzfig{../figures/appendix/e-hypergraph-example_1}
    }
    \]
\end{subfigure}
\end{figure}

\begin{algorithm}
    \caption{E-graph to e-hypergraph translation}
    \begin{algorithmic}[1]
    \Function{e-class-to-HEdge}{e\_class\_id}
    \State e\_nodes $\gets$ [e\_node$_{1}$ $\ldots$ e\_node$_{n}$] \textbf{in} H.keys() \textbf{where} H[e\_node$_{i}$] == M[e\_class\_id]
    % \State \textbf{for} [$\text{e_node}_{1}, \ldots, \text{e_node}_{n}$] \textbf{in} H \textbf{where} H[$\text{e_node}_{i}$] == M[e-class-id] \texbf{do}
    \ForAll {e\_node$_{i}$ in e\_nodes}
    \State edge $\gets$ mkEdge(e\_class\_id)
    \Comment {Create a hierarchical edge that represents an e-class}
    \State \textbf{match} e\_node$_{i}$ \textbf{with}
    \State \qquad \textbf{case} f($e_1 \ldots e_{n}$) :
    \State \qquad \qquad \textbf{if} is\_singleton(e\_class\_id) \textbf{then}
    % \If{is\_singleton(e\_class\_id)}
        \State \qquad\qquad \hspace{2em} edge $\gets$ mkEdge(f)
        \Comment {Create an edge that represents the generator f $: k \to 1$}
        \State \qquad \qquad \hspace{2em} edges $\gets$ [\Call{e-class-to-HEdge}{$e_1$} $\ldots$ \Call{e-class-to-HEdge}{$e_n$}]
        \State \qquad \qquad \hspace{2em} edge.connect(edges)
        \State \qquad \qquad \hspace{2em} \Return edge
    % \Else
    \State \qquad \qquad \textbf{else}
    \State \qquad \qquad \hspace{2em} graph $\gets$ discard([e\_node$_1$$\ldots$e\_node$_{i-1}$]) $\cup$ mkEdge(f) $\cup$ discard([e\_node$_{i+1}$ $\ldots$ e\_node$_{n}$])
        \State \qquad \qquad \hspace{2em} edges $\gets$ [\Call{e-class-to-HEdge}{$e_1$} $\ldots$ \Call{e-class-to-HEdge}{$e_n$}]
        \State \qquad \qquad \hspace{2em} edge.add(graph)
        \Comment {Add \texttt{graph} as a consistent component into hierarchical edge \texttt{edge}}
        \State \qquad \qquad \hspace{2em} edge.connect(edges)
    % \EndIf
    \EndFor
    \State \Return edge
    % \Comment {There will be a single e-node in a singleton e-class}
    % \ForAll {e-node in H \textbf{where} H[e-node] == M[e-class-id]}
    %     \State \textbf{match} e-node \textbf{with}
    %     \State \quad \textbf{case} \texttt{f}: \Return (mkEdge(f))
    %     \State \quad \textbf{case} \texttt{f}($e_1$, $\ldots$, $e_{n}$) :
    %     % this is nested
    %         \State \qquad\qquad edge $\gets$ mkEdge(f)
    %         \State \qquad\qquad edge.connect([e-class-to-HEdge(e) | e $\gets$ [$e_1$, \ldots, $e_{n}$]])
    %         \State \qquad\qquad \Return edge
    % \EndFor
    % \Else
    % \State edge $\gets$ mkEdge(M[e-class-id])
    % \ForAll {e-node in H \textbf{where} H[e-node] == M[e-class-id]}
    %     \State \textbf{match} e-node \textbf{with}
    %     \State \quad \textbf{case} \texttt{f}: edge.add(f)
    %     \State \quad \textbf{case} \texttt{f}($e_1$, $\ldots$, $e_{n}$): edge.add(f)
    % \EndFor
    % \EndIf
    \EndFunction
    \State
    \State graph $\gets \varnothing$ 
    \ForAll{e\_class\_id in H.values()}
    \State graph $\gets$ graph $\cup$ \Call {e-class-to-HEdge}{e\_class\_id})

    \EndFor
    % \Function{matchPattern}{shape}
    %     \State \textbf{match} shape \textbf{with}
    %     \State \quad \textbf{case} \texttt{Circle(radius)}:
    %     \State \quad \quad \Return $\pi \cdot \texttt{radius}^2$
    %     \State \quad \textbf{case} \texttt{Rectangle(width, height)}:
    %     \State \quad \quad \Return \texttt{width} $\times$ \texttt{height}
    %     \State \quad \textbf{case} \texttt{Triangle(base, height)}:
    %     \State \quad \quad \Return $\frac{1}{2} \times \texttt{base} \times \texttt{height}$
    %     \State \quad \textbf{case} \texttt{Square(side)}:
    %     \State \quad \quad \Return \texttt{side}$^2$
    %     \State \quad \textbf{case} \texttt{\_}:
    %     \State \quad \quad \Return ``Unknown shape''
    % \EndFunction
    \end{algorithmic}
\end{algorithm}

### 2025-01-09 LICS/sections/e-hypergraphs.tex
some updates

Two extended cospans are isomorphic if there exist isomorphisms $\alpha$, $\beta$ and $\gamma$ making the following diagram commute and such that $\alpha$ and $\gamma$ preserve order within each $p_i$.
\[
\scalebox{0.75}{
    \tikzfig{../figures/combinatorial_semantics/isomorphic_e_cospans}
}
\]
Intuitively, such construction suggests that the position of the image of $n$ within $n'$ relative to the strict internal interface does not matter; and that when considering individual components of hierarchical edges with their respective portions of strict internal input and output interfaces as plain cospans of hypergraphs (by only considering top-level vertices and edges of these components) they should be isomorphic cospans of hypergraphs in the sense from~\cite{bonchi_string_2022-1}.
Examples of isomorphic and non-isomorphic extended cospans can be found in Appendix.

>>>>

\begin{definition}
\label{def:iso}
Two extended cospans are isomorphic if there exist isomorphisms $\alpha$, $\beta$ and $\gamma$ making the following diagram commute and such that $\alpha$ and $\gamma$ preserve order within each $p_i$.
\[
\scalebox{0.75}{
    \tikzfig{../figures/combinatorial_semantics/isomorphic_e_cospans}
}
\]
\end{definition}
Intuitively, such construction suggests that the position of the image of $n$ within $n'$ relative to the strict internal interface does not matter; and that when considering individual components of hierarchical edges with their respective portions of strict internal input and output interfaces as plain cospans of hypergraphs (by only considering top-level vertices and edges of these components) they should be isomorphic cospans of hypergraphs in the sense from~\cite{bonchi_string_2022-1}.
Examples of isomorphic and non-isomorphic extended cospans can be found in Appendix.

### 2025-01-09 LICS_2025/e_graphs_with_bindings/E-graphs-with-bindings.tex
some updates

@@

Each column represents a term (or term rewrite rule), the conventional e-graph representation, and the equivalent e-string diagram representation, an e-hypergraph as developed in~\cite{ghica2024equivalencehypergraphsegraphsmonoidal}. 
The initial term, $(a*2)/2$ (which corresponds to subfigure (a)), is already represented efficiently in the e-graph by sharing the node 2.
The string diagram version is slightly different than conventional e-graphs by representing the sharing of the node 2 explicitly, using a sharing (contraction) node
\scalebox{0.3}{
	\begin{tikzpicture}
		\begin{pgfonlayer}{nodelayer}
			\node [style=vertex] (0) at (1.5, 3.5) {};
			\node [style=none] (1) at (1.5, 3) {};
			\node [style=none] (2) at (1, 4) {};
			\node [style=none] (3) at (2, 4) {};
		\end{pgfonlayer}
		\begin{pgfonlayer}{edgelayer}
			\draw (0) to (1.center);
			\draw [bend right=45] (0) to (3.center);
			\draw [bend left=45] (0) to (2.center);
		\end{pgfonlayer}
	\end{tikzpicture}
}
.

@@

@@

\begin{figure*}
    \[
        % \hspace{1.3cm}
        \adjustbox{width=\textwidth}{
        \tikzfig{../../figures/categorical-semantics/e-graph-substitution}
        }
    \]
    \caption{E-graph explicit substitution example.}
    \label{fig:e-graph-substitution}
  \end{figure*}

  \begin{figure*}
    \[
        % \hspace{1.3cm}
        \adjustbox{width=\textwidth}{
        \tikzfig{../../figures/categorical-semantics/e-string-substitution-2}
        }
    \]
    \caption{E-string diagrammatic substitution example.}
    \label{fig:e-string-substitution}
  \end{figure*}

>>>>

\subsection{String diagrams}

@@

Each column represents a term (or term rewrite rule), the conventional e-graph representation, and the equivalent string diagram representation, an e-hypergraph as developed in~\cite{ghica2024equivalencehypergraphsegraphsmonoidal}. 
The initial term, $(a*2)/2$ (which corresponds to subfigure (a)), is already represented efficiently in the e-graph by sharing the node 2.
The string diagram version is slightly different than conventional e-graphs by representing the sharing of the node 2 explicitly, using a sharing (contraction) node
\scalebox{0.3}{
	\begin{tikzpicture}
		\begin{pgfonlayer}{nodelayer}
			\node [style=vertex] (0) at (1.5, 3.5) {};
			\node [style=none] (1) at (1.5, 3) {};
			\node [style=none] (2) at (1, 4) {};
			\node [style=none] (3) at (2, 4) {};
		\end{pgfonlayer}
		\begin{pgfonlayer}{edgelayer}
			\draw (0) to (1.center);
			\draw [bend right=45] (0) to (3.center);
			\draw [bend left=45] (0) to (2.center);
		\end{pgfonlayer}
	\end{tikzpicture}
}
.

@@

A typical way on encoding $\beta$-reduction in lambda-calculus in e-graphs is by explicit substitution~\cite{EggPaper,koehler2022sketchguided}: by adding a special substitution nodes and introduce rewrite rules for them.
After the reduction has been done these nodes serve no purpose and can be removed.
We show that $\beta$-reduction comes naturally as a composition when using string diagrams for enriched closed monoidal categories.
Consider an example in Figure~\ref{fig:e-graph-substitution} where successive iterations of an e-graph for a term $(\lambda x . (y + y) + x) 1$ are shown following the $\beta$-reduction.
First the subterm for an explicit substitution is introduced which is deemed equivalent to the top-level application term.
It is then rewritten using the rules for explicit substitution at the top of each column.
At the end all explicit substitution nodes can be removed.

Now consider an example from Figure~\ref{fig:e-string-substitution} which encodes the same $\beta$-reduction for the same term.
The $\lambda$-abstraction is represented by a rounded box and the bound variable is the dangling wire.
The application node is depicted as 
\adjustbox{scale=0.015}{\begin{tikzpicture}
	\begin{pgfonlayer}{nodelayer}
		\node [style=none] (0) at (1, 6.75) {};
		\node [style=none] (1) at (1.25, 7.75) {};
		\node [style=none] (2) at (2, 8) {};
		\node [style=none] (3) at (2.75, 7.75) {};
		\node [style=none] (4) at (3, 6.75) {};
		\node [style=none] (5) at (1.5, 6.75) {};
		\node [style=none] (6) at (2.5, 6.75) {};
		\node [style=none] (7) at (2, 9) {};
		\node [style=none] (8) at (1.5, 5.75) {};
		\node [style=none] (9) at (2.5, 5.75) {};
	\end{pgfonlayer}
	\begin{pgfonlayer}{edgelayer}
		\draw [style=new edge style 2] (3.center)
			 to [in=60, out=-30, looseness=0.75] (4.center)
			 to (0.center)
			 to [in=-150, out=120, looseness=0.75] (1.center)
			 to [in=-180, out=45, looseness=0.75] (2.center)
			 to [in=135, out=0, looseness=0.75] cycle;
		\draw (2.center) to (7.center);
		\draw (5.center) to (8.center);
		\draw (9.center) to (6.center);
	\end{pgfonlayer}
\end{tikzpicture}
}. 
The reduction is just the re-attachment of wires.
Note how we avoided re-attachment for subterms as they did not use the bound variable wire, unlike in the case $[x \mapsto 1](y+y)$ before.

A common solution to replace explicit variable names in e-graphs is by using De Brujin indices~\cite{koehler2022sketchguided}.
The issue with this approach is that it interferes with sharing as equivalent subterms can be structurally different.
Consider the example $\lambda f . f (\lambda x . f x)$ which is represented as $\lambda . \%0 ((\lambda . \%1 \%0) 2)$ using De Brujin indices.
Notice how two different occurrences of the same bound variable $f$ are represented differently by $\%0$ and $\%1$ which means they can not be shared in an e-graph.
A string diagrammatic representation of the same term makes it possible for $f$ to be shared as demonstrated below.

\[
\adjustbox{scale=0.6}{
    \tikzfig{figures/de-brujin-string}
}
\]

Using De Brujin indices when performing $\beta$-reduction also introduces the complexity of representing the reduction as a rewrite rule.
Special shifting operators need to be introduced on the right-hand side of a rewrite rule which makes it not merely syntactic~\cite{koehler2022sketchguided}.

@@

\begin{figure*}
    \[
        % \hspace{1.3cm}
        \adjustbox{width=\textwidth}{
        \tikzfig{figures/e-graph-substitution}
        }
    \]
    \caption{E-graph explicit substitution example.}
    \label{fig:e-graph-substitution}
  \end{figure*}

  \begin{figure*}
    \[
        % \hspace{1.3cm}
        \adjustbox{width=\textwidth}{
        \tikzfig{figures/e-string-substitution-2}
        }
    \]
    \caption{String diagrammatic substitution example.}
    \label{fig:e-string-substitution}
  \end{figure*}
